---
title: "hw04"
author: by Julie Park
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
---    

```{r, warning=FALSE, echo=FALSE, message=F}
library(tidyverse)
library(gapminder)
library(datasets)
library(DT) #For datatable()
library(knitr) #For kable()
library(sqldf) #for SQL
library(ggridges)
```

#Exercise 1: Univariate Data Reshaping.  (Univariate Option 1)

## 1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.
```{r}
t1_wide<-gapminder %>% filter(country %in% c('Romania','Canada')) %>% 
  pivot_wider(id_cols=year, names_from=country, values_from=lifeExp)
kable(t1_wide, format='markdown')
```
<br/>

## 2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.
```{r, Fig1}
f1<-t1_wide %>% ggplot(aes(Canada,Romania,color=year)) + geom_point() + geom_smooth()

suppressMessages(print(f1))
```
<br/>

## 3. Re-lengthen the data.
```{r}
t1_long<-t1_wide %>% 
  pivot_longer(cols=-year, names_to='country', values_to='lifeExp') %>% select(country, year, lifeExp) #to set the order of variables same order as the original data
datatable(t1_long)
```

<br/><br/>

# Exercise 2: Multivariate Data Reshaping (Multivariate Option 1)
## 1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}
t2_wide<-
gapminder %>% filter(country%in%c('Canada','Romania')) %>%     
          pivot_wider(id_cols=year, names_from =country,         
                     values_from=c(lifeExp,gdpPercap), names_sep='_')

datatable(t2_wide)
```


## 2. Re-lengthen the data.
```{r}
t2_long<- t2_wide %>% 
  pivot_longer(cols=-year, 
               names_to=c('.value','country'), #order of names_to matters
               values_to=c(lifeExp, gdpPercap),names_sep='_')
datatable(t2_long)
```

<br/><br/>

# Exercise 3: Table Joins

First, two data sets were imported:
```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```


## 3.1. For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.

## 3.2 Who do we have emails for, yet are not on the guestlist?

## 3.3 Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

