---
title: "hw02"
author: by Julie Park
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  github_document
---


```{r, warning=FALSE, echo=FALSE, message=F}
library(tidyverse)
library(gapminder)
library(datasets)
library(DT) #For datatable()
library(knitr) #For kable()
```

# Exercise 1: Basic dplyr
## 1.1 
Use filter() to subset the gapminder data to 3 countries of your choice in the 1970's.

```{r 1.1}
t1<-gapminder %>% 
  filter(year %in% 1970:1979 & country %in% c('Chad','Romania','Spain'))
kable(t1,format='markdown')
```

<br/>

## 1.2 
Use the pipe operator %>% to select "country" and "gdpPercap" from your filtered dataset in 1.1.

```{r 1.2}
t2<- t1 %>% select(country, gdpPercap)
kable(t2)
```

<br/>

## 1.3 
Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that's the increase in life expectancy in your tibble. 

The following steps will be taken for exercise 1.3:

1. Make sure the year is in asending order within each country, using group_by() and arrange().
2. Obtain the lifeExp of the previous year, called prev_lifeExp, using mutate().
3. Obtain the change from the previous year, using mutate()
4. Select the entries that have experienced a drop in life expectancy compared to the year before, using filter()
```{r 1.3}
t3<-
gapminder %>% group_by(country) %>% arrange(year) %>% 
  mutate(prev_lifeExp=lag(lifeExp), change=lifeExp-prev_lifeExp) %>% 
  filter(change<0)
t3
#datatable(t3)
```

<br/>

##1.4 
Filter gapminder so that it shows the max GDP per capita experienced by each country.

```{r 1.4}
t4<-
gapminder %>% group_by(country) %>% filter(gdpPercap==max(gdpPercap))
t4
#datatable(t4)
```


</br>

##1.5 
Produce a scatterplot of Canada's life expectancy vs. GDP per capita

```{r 1.5}
gapminder %>% filter(country=='Canada') %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point() +
  scale_x_log10("GDP per capita", labels = scales::dollar_format(), limits=c(10000,40000)) +
  theme_bw() +
  ylab("Life Expectancy") + ylim(c(65,85))
```

<br/><br/>

# Exercise 2. Explore individual variables with dplyr
## One categorical variable (continent) and one continuous variable (lifeExp) are chosen for data exploration.
<br/>

## 2.1 Explore a categorical variable: continent
Africa counts 37% (n=624 rows) of the data entries, followed by Asia (n=396; 23%), Europe (n=360; 21%), Americas (n=300; 18%) and Oceania (n=24; 1%).

```{r 2.1}
t21<-
gapminder %>% count(continent) %>% mutate('%'=round(100*n/sum(n))) 
kable(t21)
```

Barplots below show the relative difference visually.
```{r 2.1.2}
gapminder %>% 
  ggplot(mapping=aes(x=continent))+geom_bar()

```


<br/>

## 2.2 Explore a continuous variable: lifeExp
As shown in the histogram below, the distribution of life expectancy (lifeExp) is not normally distributed.  The median value is 60.71 years with the interquartile range of 48.20 and 70.85.  The mean value is 59.47, which is slightly lower than the median value.  

```{r 2.2}
summary(gapminder$lifeExp)
gapminder %>% ggplot(aes(x=lifeExp))+geom_histogram(binwidth=10,center=0)
```

When this continuous  variable is categorized into 10 years, 70-80 years are the most frequent interval of life expectancy of a country.

```{r 4 2.2b}
t42<-table(cut(gapminder$lifeExp,breaks=seq(20,90,10)))
kable(t(t42))
```

<br/>

## 2.3 Explore lifeExp by continent

The distribution of lifeExp by continent is quite different as shown in the table below.  
```{r}
t23<-
gapminder %>% group_by(continent) %>% 
  summarize(Min=min(lifeExp), Q1=quantile(lifeExp,0.25), Median=median(lifeExp), Mean=mean(lifeExp), Q3=quantile(lifeExp,0.75), Max=max(lifeExp) )
kable(t23)
```
The difference in life expectancy by contient is easier to see in the graph below:
```{r}
gapminder %>% ggplot()+geom_freqpoly(aes(x=lifeExp, y=..density..,color=continent), binwidth=10,center=0)
```




<br/><br/>

# Exercise 3. Explore various plot types
## 3.1 A scatterplot of two quantitative variables

```{r 3.1}
ggplot(OrchardSprays,aes(rowpos,decrease))+geom_point()

```

## 3.2 A box plot 
Since rowpos is an integer, and takes distinct values, it can be considered as a categorical variable; a boxplot is created.
```{r 3.2}
OrchardSprays %>% 
ggplot()+geom_boxplot(aes(rowpos,decrease,group=cut_width(rowpos,1)))

```


