---
title: "hw03"
author: by Julie Park
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    keep_md: true
---


```{r, warning=FALSE, echo=FALSE, message=F}
library(tidyverse)
library(gapminder)
library(datasets)
library(DT) #For datatable()
library(knitr) #For kable()
library(sqldf) #for SQL
```

# Task Option 1
Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

__Steps taken:__

1. Perform the exploratory analysis (EDA) to determine which benchmark to use.
2. Manipulate data to create lifeExp2 (i.e. below/above benchmark)
3. Generate figures to show the results.

### 1.1 EDA
Based on the plot below, we can tell (i) the life expectancy has increased over time, with a fairly similar trajectory across all continents except for Africa, where the life expectancy plateaued around 1990s, and (ii) the life expectancy differed quite a bit by continent.  
```{r option1.1}
p1<-gapminder %>% ggplot(aes(year,lifeExp)) + geom_jitter(aes(color=continent)) + geom_smooth(aes(color=continent))

suppressMessages(print(p1))
```

Therefore, it would be reasonable to choose a benchmark that differed over time.  The remaining questions were (i) which statistic (e.g. mean vs. mean) should be chosen as a benchmark, and (ii) how often a benchmark should be updated (e.g. annual benchmark or a benchmark by a decade)?  Annual benchmark was not favoured because choosing a benchmark based on data from a single year was subject to a high variability.  A bigger bin (e.g. 10 or 15 years) was considered, instead.

In order to answer these questions, the following histograms were generated.
```{r option1.2.hist, warning=F, message=F}
gapminder %>% filter(year==1952) %>% ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For year 1952')
gapminder %>% filter(year%in%1952:1957) %>%  ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For years 1952 & 1957')
gapminder %>% filter(year%in%1952:1962) %>%  ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For years 1952, 1957 & 1962')

#gapminder %>% filter(year%in%1952:1957 & continent!='Africa') %>%  ggplot(aes(lifeExp))+geom_histogram()
```

The histograms showed that the life expectancy had a bi-modal distribution (i.e. mean is not a good choice).  The first hump was mainly due to the countries in Africa (graph not shown).  As more years of data were added, the second hump became more pronounced. A decade seemed to be a good starting interval of choice.  Therefore, a median (50th percentile) by decade was selected as a benchmark.  Note that the same benchmark would be applied across all continents.  Also, even of a benchmark value was chosen to remain constant within each decade, the boolean value (above/below benchmark) was created for each year (i.e each row of the original data).


The actual values of these benchmarks were summarized in the table below:
```{r}
bm<-gapminder %>% mutate(Decade=paste(signif(year,digit=3),'s',sep='')) %>% group_by(Decade) %>% summarize(Benchmark=median(lifeExp))
kable(bm)
```

<br/> 

### 1.2 Data manipulation
Now the benchmark is chosen, the life expectancy can be binarized (i.e. above/below the benchmark).  However, first, the benchmark value needed to be added back to the orignal data, using SQL.

```{r}
tmp<-gapminder %>% mutate(Decade=paste(signif(year,digit=3),'s',sep='')) 
gapminder1<-as_tibble(
            sqldf('select a.*,b.Benchmark from tmp as a left join 
                          bm as b on a.Decade=b.Decade'))
kable(gapminder1[1:10,])
```

Then a new variable called lifeExp2 (i.e. above/below the benchmark) was created first to create 2 additional variables: N_lt_benchmark for counts and Percent_lt_benchmark for % of N_lt_benchmark within each continent per year.
```{r}
out1<-
gapminder1 %>% mutate(lifeExp2=lifeExp<Benchmark) %>% group_by(continent,year) %>% summarize(N_lt_benchmark=sum(lifeExp2), N_total=n(), Percent_lt_benchmark=round(100*mean(lifeExp2),2)) 
datatable(out1)
```

<br/> 

### 1.3 Visualization
The first graph showed the actual number of countries that fell below the benchmark by contient.  However, a more informative summary statistic would be the percentage of countries within each continent that fell below the benchmark (2nd figure).
```{r option1.3}
out1 %>% ggplot(aes(year,N_lt_benchmark,color=continent))+geom_line()+ labs(title='Number of countries below the benchmark life expectancy over time', x='Year',y='# below benchmark')
out1 %>% ggplot(aes(year,Percent_lt_benchmark,color=continent))+geom_line()+labs(title='% of countries below the benchmark life expectancy over time', x='Year',y='# below benchmark')
```



<br/> <br/>

# Task Option 2. Get the maximum and minimum of GDP per capita for all continents.

The crude look at the data:
```{r}
out2<-gapminder %>% group_by(continent) %>% summarize(Maximum=max(gdpPercap), Minimum=min(gdpPercap))
kable(out2)
```

However, what about temporal changes?  Do we want to report maximum and minimum only by continent or also by year?
```{r option2,message=F, warning=F}
p2<-
gapminder %>% ggplot(aes(year,gdpPercap,color=continent))+ geom_jitter() + geom_smooth() + labs(title='Trajectory of GDP per capita by contient')

suppressMessages(print(p2))
```

There has been an increase in GDP per capita over the years; therefore, it would be more informative to report GDP per capita by continent and year.  In addition, a couple of extreme outliers were noted, all belonging to Asia.  A closer look at these extreme values (table below) showed that they were entries from Kuwait, from 1952 to 1977.
```{r}
kable(gapminder %>% filter(gdpPercap>59000))
```

The minimum/maximum of GDP per Capita could be summarized by continent and by year.

```{r}
out21<-gapminder %>% group_by(continent,year) %>% summarize(Maximum=max(gdpPercap), Minimum=min(gdpPercap))
datatable(out21)


```


<br/> <br/>

# Task Option 3. Look at the spread of GDP per capita within the continents.

A crude look at the distribution of GDP per capita:
```{r}
out3<-
gapminder %>% group_by(continent) %>% summarize(Min=min(gdpPercap), Mean=mean(gdpPercap), Median=median(gdpPercap), Max=max(gdpPercap))
kable(out3)
```

The graph overlaying the continents was produced in the section above (Task Option 2).  Now a seprate plot by continent could be shown for more details.

```{r option3.1, fig.width = 13, fig.height = 3}
gapminder %>% ggplot(aes(year,gdpPercap))+ geom_jitter()+ facet_wrap(~continent,nrow=1)+ labs(title='Scatter plot of GDP per capita over time by continent')
```

The figure below showed a trajectory of GDP per capita for each country within a continent for more information.
```{r option3.2, fig.width = 13, fig.height = 3}
gapminder %>% ggplot(aes(year,gdpPercap)) + geom_line(aes(group=country), alpha=0.3)+ facet_wrap(~continent,nrow=1) + labs(title='Trajectory of GDP per capita for each country by continent')
```

