---
title: "hw03"
author: by Julie Park
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
---    


```{r, warning=FALSE, echo=FALSE, message=F}
library(tidyverse)
library(gapminder)
library(datasets)
library(DT) #For datatable()
library(knitr) #For kable()
library(sqldf) #for SQL
library(ggridges)
```

# Task Option 1
Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

__Steps taken:__

1. Perform the exploratory data analysis (EDA) to determine which benchmark to use.
2. Manipulate data to create lifeExp2 (i.e. below/above benchmark) and summarize data at the continent level: number of countries below the benchnark (N_lt_benchmark) and percentage of countries below the benchnark (Percent_lt_benchmark) within each continent, given a year.
3. Generate figures to show the results.

### 1.1 EDA
Based on the plot below, we can tell (i) the life expectancy has increased over time, with a fairly similar trajectory across all continents except for Africa, where the life expectancy plateaued around 1990s, and (ii) the life expectancy differed quite a bit by continent.  
```{r option1.1}
p1<-gapminder %>% ggplot(aes(year,lifeExp)) + geom_jitter(aes(color=continent)) + geom_smooth(aes(color=continent))

suppressMessages(print(p1))
```

Therefore, it would be reasonable to choose a benchmark that differed over time.  The remaining questions were (i) which statistic (e.g. mean vs. mean) should be chosen as a benchmark, and (ii) how often a benchmark should be updated (e.g. annual benchmark or a benchmark by a decade)?  Annual benchmark was not favoured because choosing a benchmark based on data from a single year was subject to a high variability.  A bigger bin (e.g. 10 or 15 years) was considered, instead.

In order to answer these questions, the following histograms were generated.
```{r option1.2.hist, warning=F, message=F, figure.width=5, figure.height=2}
gapminder %>% filter(year==1952) %>% ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For year 1952')
gapminder %>% filter(year%in%1952:1957) %>%  ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For years 1952 & 1957')
gapminder %>% filter(year%in%1952:1962) %>%  ggplot(aes(lifeExp))+geom_histogram(bins=50) + labs(title='For years 1952, 1957 & 1962')

#gapminder %>% filter(year%in%1952:1957 & continent!='Africa') %>%  ggplot(aes(lifeExp))+geom_histogram()
```

The histograms showed that the life expectancy had a bi-modal distribution (i.e. mean is not a good choice).  The first hump was mainly due to the countries in Africa (graph not shown).  As more years of data were added, the second hump became more pronounced. A decade seemed to be a good starting interval of choice.  Therefore, a median (50th percentile) by decade was selected as a benchmark.  Note that the same benchmark would be applied across all continents.  Also, even of a benchmark value was chosen to remain constant within each decade, the boolean value (above/below benchmark) was created for each year (i.e each row of the original data).


The actual values of these benchmarks were summarized in the table below:
```{r}
bm<-gapminder %>% mutate(Decade=paste(signif(year,digit=3),'s',sep='')) %>% group_by(Decade) %>% summarize(Benchmark=median(lifeExp))
kable(bm, format='markdown')
```

<br/> 

### 1.2 Data manipulation
Now the benchmark is chosen, the life expectancy can be binarized (i.e. above/below the benchmark).  However, first, the benchmark value needed to be added back to the orignal data, using SQL.

```{r}
tmp<-gapminder %>% mutate(Decade=paste((year-year%%10),'s',sep='')) 
gapminder1<-as_tibble(
            sqldf('select a.*,b.Benchmark from tmp as a left join 
                          bm as b on a.Decade=b.Decade'))
kable(gapminder1[1:10,], format='markdown')
```

Then a new variable called lifeExp2 (i.e. above/below the benchmark) was created to summarize data by continent and by year: number of countries below the benchnark (N_lt_benchmark) and percentage of countries below the benchnark (Percent_lt_benchmark) within each continent in a given year.
```{r}
out1<-
gapminder1 %>% mutate(lifeExp2=lifeExp<Benchmark) %>% group_by(continent,year) %>% summarize(N_lt_benchmark=sum(lifeExp2), N_total=n(), Percent_lt_benchmark=round(100*mean(lifeExp2),2)) 
datatable(out1)
```

<br/> 

### 1.3 Visualization
The first graph showed the actual number of countries that fell below the benchmark by continent.  However, a more informative summary statistic would be the percentage of countries within each continent that fell below the benchmark (2nd figure).
```{r option1.3}
out1 %>% ggplot(aes(year,N_lt_benchmark,color=continent))+geom_line()+ labs(title='Number of countries below the benchmark life expectancy over time', x='Year',y='# below benchmark')
out1 %>% ggplot(aes(year,Percent_lt_benchmark,color=continent))+geom_line()+labs(title='% of countries below the benchmark life expectancy over time', x='Year',y='# below benchmark')
```



<br/> <br/>

# Task Option 2. Get the maximum and minimum of GDP per capita for all continents.

The overall maximum and minimum of GDP per capital across all continents:
```{r}
out20<-as.array(summary(gapminder$gdpPercap)[c('Max.','Min.')])
kable(t(out20))
```

<br/>
The maximum and minimum by continent were summarized in the table below:
```{r}
out2<-gapminder %>% group_by(continent) %>% summarize(Maximum=max(gdpPercap), Minimum=min(gdpPercap))
kable(out2, format='markdown')
```
So now we know the global minimum of GDP per capita belonged to Africa, and the global maximum to Asia.

Then, what about the temporal changes?  Do we want to report maximum and minimum only by continent or also by year?
```{r option2,message=F, warning=F}
p2<-
gapminder %>% ggplot(aes(year,gdpPercap,color=continent))+ geom_jitter() + geom_smooth() + labs(title='Trajectory of GDP per capita by continent')

suppressMessages(print(p2))
```

There has been an increase in GDP per capita over the years; therefore, it would be more informative to report GDP per capita by continent and year.  In addition, a couple of extreme outliers were noted, all belonging to Asia.  A closer look at these extreme values (table below) showed that they were entries from Kuwait, from 1952 to 1977.
```{r}
kable(gapminder %>% filter(gdpPercap>59000), format='markdown')
```

<br/>
The maximum and minimum of GDP per Capita were summarized by continent and by year.

```{r}
out21<-gapminder %>% group_by(continent,year) %>% summarize(Maximum=max(gdpPercap), Minimum=min(gdpPercap))
datatable(out21)


```


<br/> <br/>

# Task Option 3. Look at the spread of GDP per capita within the continents.

The crude look at the data:
```{r}
gapminder %>% ggplot(aes(gdpPercap,continent))+geom_density_ridges()+geom_point(alpha=0.2) 
```

It looked like Europe, Oceania and Asia had much more spread of data than Americas and Afria.  

<br/>
To quantify the spread of GDP per capita, the summary statistics table was created below:
```{r}
out3<-
gapminder %>% group_by(continent) %>% summarize(Min=min(gdpPercap), Mean=mean(gdpPercap), Q1=quantile(gdpPercap,0.25), Median=median(gdpPercap), Q3=quantile(gdpPercap, 0.75), Max=max(gdpPercap))
kable(out3, format='markdown')
```

Boxplots are always more informative than a summary table above:
```{r options3.1}
gapminder %>% ggplot(aes(continent,gdpPercap))+geom_boxplot() +
  labs(title='Boxplot of GDP per capita by continent (including all entries)')
```

In the earlier section, it was shown that the outliers in Asia belonged to Kuwait, from 1952 to 1977.  The figure can be zoomed in by excluding these outliers.  Note that excluding 6 extreme points in Asia did not alter the median and interquartile range of Asia significantly (data not shown).
```{r option3.2}
gapminder %>% filter(!(country=='Kuwait' & year%in%1952:1977)) %>%  ggplot(aes(continent,gdpPercap))+geom_boxplot() + labs(title='Boxplot of GDP per capita by continent (excluding extreme outliers)')
```

Now it is easier to see the difference in the median and the interquartile range (i.e. hinges of a box) as well as the outliers by continent.